---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "~/dropbox/github/csslab/lab05/")
```


# Data

116th Congress (2019-2021)

106th Congress (1999-2001)

96th Congress (1979-1981)

\url{https://voteview.com/data}

Note `party_code` of 100 is Dem, 200 is Rep, and 328 is Ind.

```{r, fig.width=6, fig.height=5}
x <- read.csv("HS096_members.csv")

r <- rgb(red = 1, green = 0, blue = 0, alpha = .5, maxColorValue = 1)
b <- rgb(red = 0, green = 0, blue = 1, alpha = .5, maxColorValue = 1)
g <- rgb(red = 0, green = 1, blue = 0, alpha = .75, maxColorValue = 1)

plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = ifelse(x$party_code == 100, r,
                  ifelse(x$party_code == 200, b, g)),
     pch = ifelse(x$chamber == "House", 16, 17),
     cex = 1)
```

# kmeans

- Find $k$ clusters with hard assignments


# Optimization algorithm

What are the steps?

1. Choose random centroids

Do until convergence:

- Update $\tau_i$ -- find closest centroid for each observation

- Update $\theta_k$ -- calculate new centroids

- Update change in objective function


# Let's do one iteration

- plot each step

- try to make code generalizable to any $K$ and any dimension of $X$

```{r}
x <- x[,c("nominate_dim1", "nominate_dim2")]
```

```{r}
set.seed(43525)
K <- 4
centers <- matrix(NA, nrow = K, ncol = ncol(x))
for(i in 1:ncol(centers)){
    centers[,i] <- runif(K, min = min(x[,i]), max = max(x[,i]))
}
colnames(centers) <- colnames(x)
rownames(centers) <- paste0("center", 1:K)
centers
```


```{r}
plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = "black",
     cex = 1)
points(centers, pch = "+", cex = 3, col = 1:K)
```

```{r}
new_assign <- rep(NA, nrow(x))
for(i in 1:nrow(x)){
    x_i <- x[i, ]
    dists <- rep(NA, K)
    for(j in 1:K){
        dists[j] <- dist(rbind(x_i, centers[j,]))
    }
    new_assign[i] <- which.min(dists)    
}
```


```{r}
plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = new_assign,
     cex = 1)
points(centers, pch = "+", cex = 3, col = 1:K)
```

```{r}
new_centers <- matrix(NA, nrow = K, ncol = ncol(x))
for(i in 1:nrow(new_centers)){
    new_centers[i,] <- colMeans(x[new_assign == i, ])
}
```

```{r}
plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = new_assign,
     cex = 1)
points(new_centers, pch = "+", cex = 3, col = 1:K)
```

# Now iterate until convergence


```{r}
calc_loss <- function(x, K, centers){
    loss <- 0
    for(i in 1:nrow(x)){
        x_i <- x[i, ]
        dists <- rep(NA, K)
        for(j in 1:K){
            dists[j] <- dist(rbind(x_i, centers[j,]))
        }
        loss <- loss + min(dists)^2
    } 
    return(loss)
}


my_kmeans <- function(x, K, max_iter, eps, seed){
    set.seed(seed)
    # initialize random centroids
    centers <- matrix(NA, nrow = K, ncol = ncol(x))
    for(i in 1:ncol(centers)){
        centers[,i] <- runif(K, min = min(x[,i]), max = max(x[,i]))
    }
    # initial loss
    # uses t-1 centers and assignments given those centers
    old_loss <- calc_loss(x = x, K = K, centers = centers)
    
    change <- 2*eps
    iter <- 0
    while(change > eps & iter < max_iter){
        
        # assign each point to closest centroid
        assign <- rep(NA, nrow(x))
        for(i in 1:nrow(x)){
            x_i <- x[i, ]
            dists <- rep(NA, K)
            for(j in 1:K){
                dists[j] <- dist(rbind(x_i, centers[j,]))
            }
            assign[i] <- which.min(dists)
        }
        
        # calculate new centroids
        centers <- matrix(NA, nrow = K, ncol = ncol(x))
        for(i in 1:nrow(centers)){
            centers[i,] <- colMeans(x[assign == i, ])
        }
        
        # update change in loss
        new_loss <- calc_loss(x = x, K = K, centers = centers)
        change <- old_loss - new_loss
        old_loss <- new_loss
        
        iter <- iter + 1
    }

    return(list(assignments = assign, centers = centers, n_iter = iter, obj_func = old_loss))
}


out1 <- my_kmeans(x = x, K = 3, max_iter = 100, eps = .001, seed = 1234)
out2 <- my_kmeans(x = x, K = 3, max_iter = 100, eps = .001, seed = 9873)
```

```{r, fig.height=6, fig.width=4}
par(mfrow = c(2,1))

plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = out1$assignments,
     cex = 1)
points(out1$centers, pch = "+", cex = 3, col = 1:K)
mtext(text = paste("Obj Func:", out1$obj_func))

plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = out2$assignments,
     cex = 1)
points(out2$centers, pch = "+", cex = 3, col = 1:K)
mtext(text = paste("Obj Func:", out2$obj_func))
```
```

