---
title: Kmeans
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "~/dropbox/github/csslab/lab05/")
```


# Data

116th Congress (2019-2021)

106th Congress (1999-2001)

96th Congress (1979-1981)

\url{https://voteview.com/data}

Note `party_code` of 100 is Dem, 200 is Rep, and 328 is Ind.

```{r, fig.width=6, fig.height=5}
x <- read.csv("HS096_members.csv")

r <- rgb(red = 1, green = 0, blue = 0, alpha = .5, maxColorValue = 1)
b <- rgb(red = 0, green = 0, blue = 1, alpha = .5, maxColorValue = 1)
g <- rgb(red = 0, green = 1, blue = 0, alpha = .75, maxColorValue = 1)

plot(x = x$nominate_dim1, y = x$nominate_dim2,
     col = ifelse(x$party_code == 100, r,
                  ifelse(x$party_code == 200, b, g)),
     pch = ifelse(x$chamber == "House", 16, 17),
     cex = 1)
```

# kmeans

- Find $k$ clusters with hard assignments


# Optimization algorithm

What are the steps?

1. Choose random centroids

2. Do until convergence:

    - Update $\tau_i$ -- find closest centroid for each observation

    - Update $\theta_k$ -- calculate new centroids

    - Update change in objective function


# Let's code one iteration

- plot each step

- try to make code generalizable to any $K$ and any dimension of $X$

```{r}
x <- x[,c("nominate_dim1", "nominate_dim2")]
```



# Now adapt code to iterate until convergence



